Complete Real-Time Liver Model Including Glisson's Capsule, Vascularization and Parenchyma

### 2013-03-02 Doc/Miccai2013/miccai2013.tex
ADD: small changes in the text
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@551 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

Rather a small number of studies have been conducted on the last liver constituent, the  Glisson's capsule.
Quantitative results of experiments on a porcine liver have been published in~\cite{Umale2013}; the measurements indicate that although being very 
thin (10--20$\mu m$), the capsule shows to be stiff in tensile tests: the Young's modulus of the capsule (reported as $8.22\pm3.42$\,MPa) 
exceeds the values for the parenchyma by three orders of magnitude.
This suggests that the mechanical influence of the membrane on the liver behaviour is not negligible.

@@

\subsection{Glisson's Capsule} %{{{
\label{ss:capsuleModel}
The thickness of the Glisson's capsule is relatively small. Umale et al.
\cite{Umale2011} reported for porcine liver values in range of 10-20
$\mu$m.
It is not possible to model such thin structure with classical tetrahedral
elements. Modeling this non-homogeneous volumetric object would require a
dense mesh to avoid numerical instabilities and would significantly
violate the speed requirements imposed on medical simulators.
Instead modeling the capsule with two-dimensional elements that abstract from the
thickness in the third dimension seems
as a natural choice. Mechanics provides this FE tool in the form of membranes
and shells. Based on the small thickness (compared to the surface area) we also
assume negligible bending forces and propose a model based on membrane
elements. 
To maintain simplicity of the composite model we choose simple triangular
elements with constant strain (CST).

The computation of elastic stiffness matrix follows the common derivation
%
\begin{eqnarray}
  \Mat{K}^m & = & \int_V \Mat{B}^T \Mat{E} \Mat{B} dV     \label{mem1} \\
            & = & h \int_A \Mat{B}^T \Mat{E} \Mat{B} dA   \label{mem2} \\
            & = & h A \Mat{B}^T \Mat{E} \Mat{B}           \label{mem3}
\end{eqnarray}
%
\noindent%
where $\Mat{B}$ is the strain-displacement matrix, $\Mat{E}$ the material
matrix, $h$ is the thickness and $A$ area of the element. In the previous
\eqref{mem2} follows from the fact that we assume constant thickness of the
element and \eqref{mem3} follows from the fact that the strain-displacement
matrix is in our case constant. The strain-displacement matrix for the CST
element can be expressed as:
%
\begin{equation}
  \Mat{B} = \frac{1}{2A} \begin{bmatrix}
    y_{23} & 0      & y_{31} & 0      & y_{12} & 0 \\
         0 & x_{32} & 0      & x_{13} & 0      & x_{21} \\
    x_{32} & y_{23} & x_{13} & y_{31} & x_{21} & y_{12}
  \end{bmatrix}
\end{equation}
%
The values $x_{ij} = x_i - x_j$ and $y_{ij} = y_i - y_j$ are computed from
the $x$ or $y$ coordinates of the nodes $i,j$ of the triangular element.
The reader can reffer to the respective literature for more thorough
description \cite{Felippa2003}.

We again use linear elastic material and employ the corotational formulation
for the triangular elements.

@@

\subsection{Coupling Between Capsule and Parenchyma} %{{{

The literature reports high cohesion between capsule and parenchyma.
Based on this
property we may assume no relative motion of the capsule to the parenchyma.
While in general we could use arbitrary surface mesh for a capsule we use
the fact that the parenchyma is modeled by tetrahedral elements which have
triangular faces. Because the boundary of the volumetric mesh is already
triangulated, we only extract the boundary triangles which we use for the
model of capsule.

Using directly the boundary of the tetrahedral mesh does not only solve the
problem of obtaining the surface mesh, but has one more advantage. Nodes
of the triangular mesh overlap with nodes of the tetrahedral mesh and we do
not have to project nodes of triangular mesh onto the tetrahedra. More over,
the stiffness matrices for capsule and parenchyma are easily assembled
together.% and solved as one system.
%
%\CD{The following may appear as "trivial"... maybe we can remove this part at the end if we need space} 
%Without the loss of generality we can assume the tetrahedron consists of
%nodes $p_1, p_2, p_3$ and $p_4$ and the boundary triangle has nodes $p_1, p_2$
%and $p_3$. We can reorder the degrees of freedom so that the stiffness
%matrix $\Mat{K}^t$ for the tetrahedron can be written as:
%%
%\begin{equation}
%  \Mat{K}^t = \left[\begin{array}{c|c}
%      \Mat{K}^t_{1-3,1-3} & \Mat{K}^t_{1-3,4} \\
%      \hline
%      \Mat{K}^t_{4,1-3} & \Mat{K}^t_{4,4} \\
%  \end{array}\right]
%\end{equation}
%%
%Then the assembled stiffness matrix for the element is:
%%
%\begin{equation}
%  \Mat{K} = \left[\begin{array}{c|c}
%      \Mat{K}^t_{1-3,1-3} & \Mat{K}^t_{1-3,4} \\
%      \hline
%      \Mat{K}^t_{4,1-3} & \Mat{K}^t_{4,4} \\
%  \end{array}\right]
%  +
%  \left[\begin{array}{c|c}
%      \Mat{K}^m & 0 \\
%      \hline
%      0 & 0 \\
%  \end{array}\right]
%\end{equation}
%%
%Where $\Mat{K}^m$ is the stiffness matrix of the triangular membrane.
The resulting
system of linear equations is then solved by direct LDL solver.

@@

\section{Conclusion} %{{{

A composite model for simulation of complete liver is presented in the
paper. The work of Peterlik et al. for vaskularized liver model was
extended by addig the capsule modeled with membrane FEM model mechanically
coupled with the parenchyma. To validate the model a simulation of local
deformations was performed and was found in an aggrement with the results
measured and published by Hollenstein et al.

While the influence of the capsule on local deformations was previously
studied in the literature, it's significance on the global scale
deformations of the liver remained unexplored. We have performed a set of
natural global deformations of the complete liver and shown that the
capsule, despite its small thickness, plays a significant role also in
global context.

>>>>

Rather a small number of studies have been conducted on the last liver constituent, the  Glisson's capsule.
Quantitative results of experiments on a porcine liver have been published in~\cite{Umale2011}; the measurements indicate that although being very 
thin (10--20$\mu m$), the capsule shows to be stiff in tensile tests: the Young's modulus of the capsule (reported as $8.22\pm3.42$\,MPa) 
exceeds the values for the parenchyma by three orders of magnitude.
This suggests that the mechanical influence of the membrane on the liver behaviour is not negligible.

@@

\subsection{Glisson's Capsule} %{{{
\label{ss:capsuleModel}
The thickness of the Glisson's capsule is relatively small: values in range of 10-20
$\mu$m have been reported in~\cite{Umale2011}.
It is not possible to model such thin structure with classical tetrahedral
elements, if real-time aspect of the simulation is to be preserved.
Furthermore, modeling both the tissue and the capsule would require an extremely 
dense mesh to avoid numerical instabilities and thus would significantly
violate the speed requirements required for medical simulators.
Instead, modeling the capsule with two-dimensional elements that abstract from the
thickness in the third dimension seems
to be a natural choice. In the elasticity theory, this functionality is usually provided by membrane and shell elements.
Based on the small thickness (compared to the surface area) we also
assume negligible bending forces and propose a model based on membrane
elements. 
To maintain simplicity of the composite model we choose simple triangular
elements with constant strain (CST).

The computation of elastic stiffness matrix follows the common derivation
%
\begin{eqnarray}
  \Mat{K}^m & = & \int_V \Mat{B}^T \Mat{E} \Mat{B} dV     \label{mem1} \\
            & = & h \int_A \Mat{B}^T \Mat{E} \Mat{B} dA   \label{mem2} \\
            & = & h A \Mat{B}^T \Mat{E} \Mat{B}           \label{mem3}
\end{eqnarray}
%
where $\Mat{B}$ is the strain-displacement matrix, $\Mat{E}$ the material
matrix, $h$ is the thickness and $A$ area of the element. In the previous
\eqref{mem2} follows from the fact that we assume constant thickness of the
element and \eqref{mem3} follows from the fact that the strain-displacement
matrix is constant in our case. The strain-displacement matrix for the CST
element can be expressed as:
%
\begin{equation}
  \Mat{B} = \frac{1}{2A} \begin{bmatrix}
    y_{23} & 0      & y_{31} & 0      & y_{12} & 0 \\
         0 & x_{32} & 0      & x_{13} & 0      & x_{21} \\
    x_{32} & y_{23} & x_{13} & y_{31} & x_{21} & y_{12}
  \end{bmatrix}
\end{equation}
%
The values $x_{ij} = x_i - x_j$ and $y_{ij} = y_i - y_j$ are computed from
the $x$ or $y$ coordinates of the nodes $i,j$ of the triangular element.
The reader can refer to the respective literature for more thorough
description~\cite{Felippa2003}.

We use linear elastic material and employ the corotational formulation
for the triangular elements.

@@

\subsection{Coupling Between Capsule and Parenchyma} %{{{
The literature reports high cohesion between capsule and parenchyma.
Based on this property we assume there is no relative motion of the capsule \wrt\ the parenchyma.
Although an arbitrary surface mesh could be used to model the capsule, we exploit 
the fact that the parenchyma is modeled by tetrahedral elements having
triangular faces. Thus, as the boundary of the volumetric mesh is already
triangulated, we simply employ the triangles on the mesh surface to model the capsule.

Using directly the boundary of the tetrahedral mesh does not only solve the
problem of building the surface mesh, but has one more advantage: the nodes
of the triangular mesh coincide with the nodes of the tetrahedral mesh, so no projection of one mesh onto the other is needed.
Moreover, the stiffness matrices for capsule and parenchyma are then easily assembled
together.% and solved as one system.
%
%\CD{The following may appear as "trivial"... maybe we can remove this part at the end if we need space} 
%Without the loss of generality we can assume the tetrahedron consists of
%nodes $p_1, p_2, p_3$ and $p_4$ and the boundary triangle has nodes $p_1, p_2$
%and $p_3$. We can reorder the degrees of freedom so that the stiffness
%matrix $\Mat{K}^t$ for the tetrahedron can be written as:
%%
%\begin{equation}
%  \Mat{K}^t = \left[\begin{array}{c|c}
%      \Mat{K}^t_{1-3,1-3} & \Mat{K}^t_{1-3,4} \\
%      \hline
%      \Mat{K}^t_{4,1-3} & \Mat{K}^t_{4,4} \\
%  \end{array}\right]
%\end{equation}
%%
%Then the assembled stiffness matrix for the element is:
%%
%\begin{equation}
%  \Mat{K} = \left[\begin{array}{c|c}
%      \Mat{K}^t_{1-3,1-3} & \Mat{K}^t_{1-3,4} \\
%      \hline
%      \Mat{K}^t_{4,1-3} & \Mat{K}^t_{4,4} \\
%  \end{array}\right]
%  +
%  \left[\begin{array}{c|c}
%      \Mat{K}^m & 0 \\
%      \hline
%      0 & 0 \\
%  \end{array}\right]
%\end{equation}
%%
%Where $\Mat{K}^m$ is the stiffness matrix of the triangular membrane.
The resulting system of linear equations is solved by direct solver based on Cholesky decomposition.

@@

\section{Conclusion} %{{{
In the paper, a complete model of the liver was presented taking into account 
three constituents of the organ, each being modelled with a different type 
of finite elements: tetrahedral corotational model used the parenchyma, serially 
linked corotational beams based on Tymoshenko formulation modelling the vascularization 
and finally triangular elements with constant strain employed for the capsule. 
The model was first validated for local displacements: the aspiration test presented in~\cite{Hollenstein2006} 
was succesfully reproduced, showing a good agreement between the simulation and experiments. 
Further, the model was used to demonstrate global influence of the capsule when the organ 
undergoes large deformations.

In the future work, we plan to employ the model to model other organs where the capsule play an important role
(for example kidneys). We will also focus on better understanding the interaction between the organs in the abdominal 
cavity.

%While the influence of the capsule on local deformations was previously
%studied in the literature, it's significance on the global scale
%deformations of the liver remained unexplored. We have performed a set of
%natural global deformations of the complete liver and shown that the
%capsule, despite its small thickness, plays a significant role also in
%global context.
%
%}}}
